Research Notes

### 2018-04-07 writups/notes.tex
implement radiative conditions, writeup first of nonlinear

To get an intuition for how Dedalus and fluid mechanics works, we will solve
some toy problems. Recall fluid equations in the presence of a uniform
gravitational field $\vec{g} = -g\hat{z}$:
\begin{equation}
    \begin{split}
        \rd{\rho}{t} + \vec{\nabla} \cdot \vec{u} &= 0,\\
        \rd{\vec{u}}{t} + \frac{\vec{\nabla}P}{\rho} - \vec{g} &= 0.
    \end{split}\label{eq:fluid_eq}
\end{equation}

@@

Let's restore the $\rho_1 g$ term now. For funsies, we begin by solving for
arbitrary stratification $\rho_0(z)$ first. The fluid equations to first order
reduce to
\begin{equation}
    \begin{split}
        \pd{\rho_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\rho_0 &= 0,\\
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} &= -\frac{\vec{\nabla}P_1}{\rho_0}
            - \frac{\rho_1 g}{\rho_0}
    \end{split}\label{eq:lin_incomp}
\end{equation}
We expect there to be some $z$ dependence in the amplitude, so we substitute
variables of form $e^{i(kx - \omega t)}$ and do not specify the $z$ dependence.
This gives us
\begin{equation}
    \begin{split}
        -i\omega \rho_1 - u_{1z}\pd{\rho_0}{z} &= 0,\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1 g}{\rho_g} &= 0.
    \end{split}
\end{equation}
We substitute $N^2 = -\frac{g}{\rho_0}\pd{\rho_0}{z}$ to obtain
\begin{subequations}
    \begin{align}
        -i\omega \rho_1 - u_{1z}\frac{\rho_0N^2}{g} &= 0,\label{eq:lin.cont}\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\label{eq:lin.div}\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\label{eq:lin.mom_x}\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1g}{\rho_0} &= 0.\label{eq:lin.mom_z}
    \end{align}
\end{subequations}
Eliminating $u_{1x}$ by substituting~\eqref{eq:lin.div}
into~\eqref{eq:lin.mom_x} and $\rho_1$ by substituting~\eqref{eq:lin.cont}
into~\eqref{eq:lin.mom_z} give
\begin{subequations}
    \begin{align}
        i\omega \pd{u_{1z}}{z} + \frac{k_x^2 P_1}{\rho_0} &= 0,
            \label{eq:lin.eq1}\\
        \p*{\omega^2 - N^2 }u_{1z} + \frac{i\omega}{\rho_0}\pd{P_1}{z} &= 0.
            \label{eq:lin.eq2}
    \end{align}
\end{subequations}
Finally, we multiply~\eqref{eq:lin.eq1} with $\rho_0$ and differentiate
$\mathrm{d}z$ and combine with~\eqref{eq:lin.eq2} to give
\begin{equation}
    \rtd{u_{1z}}{z} + \frac{1}{\rho_0}\pd{\rho_0}{z}\pd{u_{1z}}{z}
        + k_x^2\p*{\frac{N^2}{\omega^2} - 1}u_{1z} = 0.\label{eq:lin.incomp.gen}
\end{equation}

@@

Sourced from \url{https://people.maths.ox.ac.uk/trefethen/6all.pdf}. There are
largely two ways to do this. The first is to add a damping zone, a region in
which $\dot{q} = -q/\tau_d(z)$ where $\tau_d(0) = 0$ and increases to some
non-small number for some $z_0$ beyond which we want damping. One form, chosen
in Ryan and Dong's paper, is to use multiplicative factor $f(z) = \max\s*{0, 1 -
\frac{\p*{z - z_b}^2}{\p*{z_d - z_b}^2}}$ where $z_b$ is where one begins
supression and $z_b$ is the boundary. Then a dynamical variable $q$ can be
supressed via $\dot{q} = (\dots) - f(z)\frac{q - q_0}{\tau}$ where $\tau$ is the
dynamical timescale on which to damp and $q_0$ is the value to damp to.

@@

>>>>

To get an intuition for how Dedalus and fluid mechanics works, we will solve
some toy problems. Recall fluid equations in the presence of a uniform
gravitational field $\vec{g} = -g\hat{z}$:
\begin{equation}
    \begin{split}
        \pd{\rho}{t} + \vec{\nabla} \cdot \p*{\rho \vec{u}} &= 0,\\
        \rd{\vec{u}}{t} + \frac{\vec{\nabla}P}{\rho} - \vec{g} &= 0.
    \end{split}\label{eq:fluid_eq}
\end{equation}

@@

Let's restore the $\rho_1 g$ term now. For funsies, we begin by solving for
arbitrary stratification $\rho_0(z)$ first. The fluid equations to first order
reduce to
\begin{equation}
    \begin{split}
        \pd{\rho_1}{t} + \vec{u}_1 \cdot \p*{\vec{\nabla}\rho_0} &= 0,\\
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} &= -\frac{\vec{\nabla}P_1}{\rho_0}
            - \frac{\rho_1 g}{\rho_0}
    \end{split}\label{eq:lin_incomp}
\end{equation}
We expect there to be some $z$ dependence in the amplitude, so we substitute
variables of form $e^{i(kx - \omega t)}$ and do not specify the $z$ dependence.
This gives us
\begin{equation}
    \begin{split}
        -i\omega \rho_1 - u_{1z}\pd{\rho_0}{z} &= 0,\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1 g}{\rho_g} &= 0.
    \end{split}
\end{equation}
We substitute $N^2 = -\frac{g}{\rho_0}\pd{\rho_0}{z}$ to obtain
\begin{subequations}
    \begin{align}
        -i\omega \rho_1 - u_{1z}\frac{\rho_0N^2}{g} &= 0,\label{eq:lin.cont}\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\label{eq:lin.div}\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\label{eq:lin.mom_x}\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1g}{\rho_0} &= 0.\label{eq:lin.mom_z}
    \end{align}
\end{subequations}
Eliminating $u_{1x}$ by substituting~\eqref{eq:lin.div}
into~\eqref{eq:lin.mom_x} and $\rho_1$ by substituting~\eqref{eq:lin.cont}
into~\eqref{eq:lin.mom_z} give
\begin{subequations}
    \begin{align}
        i\omega \pd{u_{1z}}{z} + \frac{k_x^2 P_1}{\rho_0} &= 0,
            \label{eq:lin.eq1}\\
        \p*{\omega^2 - N^2 }u_{1z} + \frac{i\omega}{\rho_0}\pd{P_1}{z} &= 0.
            \label{eq:lin.eq2}
    \end{align}
\end{subequations}
Finally, we multiply~\eqref{eq:lin.eq1} with $\rho_0$ and differentiate
$\mathrm{d}z$ and combine with~\eqref{eq:lin.eq2} to give
\begin{equation}
    \rtd{u_{1z}}{z} + \frac{1}{\rho_0}\pd{\rho_0}{z}\pd{u_{1z}}{z}
        + k_x^2\p*{\frac{N^2}{\omega^2} - 1}u_{1z} = 0.\label{eq:lin.incomp.gen}
\end{equation}

@@

Sourced from \url{https://people.maths.ox.ac.uk/trefethen/6all.pdf}. There are
largely two ways to do this. The first is to add a damping zone, a region in
which $\dot{q} = -q/\tau_d(z)$ where $\tau_d(0) = 0$ and increases to some
non-small number for some $z_0$ beyond which we want damping. One form, chosen
in Ryan and Dong's paper, is to use multiplicative factor $f(z) = \max\s*{0, 1 -
\frac{\p*{z - z_b}^2}{\p*{z_d - z_b}^2}}$ where $z_b$ is where one begins
supression and $z_b$ is the boundary. Then a dynamical variable $q$ can be
supressed via $\dot{q} = (\dots) - f(z)\frac{q - q_0}{\tau}$ where $\tau$ is the
dynamical timescale on which to damp and $q_0$ is the value to damp to. We will
apply this only to the velocity variables with an eye to extending this approach
to the nonlinear regime, where the velocity variables should still be damped to
zero but $\rho, P$ must capture the stratification and so cannot easily be
artificially damped (maybe? Revisit if this does poorly compared to damping all
linear variables, and damp the nonlinear variables to their
equilibrium/stratified values).

@@

From the numerical simulations, it is clear that reflections are well supressed
initially, but the the simulation blows up. This is because our supression is
imperfect (differential approximation to a pseudodifferential operator) and
since our system is dissipation free, reflection grows. Moreover, since any
reflected components seem to have different $k_x, k_z, \omega$, the second time
they're incident on the boundary, our above condition becomes wildly inaccurate.

\section{Incompressible, Stratified w/ Gravity, Nonlinear}

We consider the problem where $\rho_1 \sim \rho_0$, where we cannot make such a
perturbative expansion. We write down thus nonlinear fluid equations
\begin{equation}
    \begin{split}
        \pd{\rho}{t} + \vec{u} \cdot \p*{\vec{\nabla}\rho} &= 0,\\
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}}{t} &= -\frac{\vec{\nabla}P}{\rho} - g\hat{z}.
    \end{split}\label{eq:nonlin_incomp}
\end{equation}
Note that we cannot even subtract off hydrostatic equilibrium anymore since
$\rho$ can deviate greatly from the normal $\rho_0e^{-z/H}$!

Since Dedalus cannot handle a field quantity in the denominator, we must get rid
of $\rho$ somehow. We choose to do this by recasting the equations in terms of
$\rho^{-1}$, which is everywhere nonzero and finite since $\rho$ is everywhere
nonzero and finite. But then $\pd{\rho}{t} = -\rho^2\pd{\rho^{-1}}{t}$ while
$\vec{\nabla}\rho = -\rho^2\vec{\nabla}\rho^{-1}$, and so factoring out the
$\rho^2$ we can simply rewrite
\begin{equation}
    \begin{split}
        \pd{\rho^{-1}}{t} + \vec{u} \cdot \p*{\vec{\nabla}\rho^{-1}} &= 0,\\
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}}{t} &= -\rho^{-1}\vec{\nabla}P - g\hat{z}.
    \end{split}\label{eq:nonlin_incomp}
\end{equation}

We must also consider how we implement the outgoing boundary condition in this
regime. If we use the radiative boundary condition formula, thankfully we have
little to change, though it also doesn't do a great job of supressing
reflections. If we instead use a sponge zone, we should be careful about how we
damp $\rho^{-1}$, since we obviously should not try damping it to zero! This is
why we initially chose to only apply our sponge zone to the velocity variables,
even in the linear problem.
